Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-02-26 doc/sfincsPaper/sfincsPaper.tex
Made notation for tensors consistent in the manuscript.
Minor modifications to paper

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\tens}[1]{\overleftrightarrow{#1}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\vi}{v_{\mathrm{i}}}
\newcommand{\me}{m_{\mathrm{e}}}
\newcommand{\mi}{m_{\mathrm{i}}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{\mathrm{ii}}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\Erres}{E_r^{\mathrm{res}}}
\newcommand{\Bpol}{B_{\mathrm{pol}}}
\newcommand{\Sap}{S_{a\mathrm{p}}}
\newcommand{\Sah}{S_{a\mathrm{h}}}
%\newcommand{\changed}[1]{#1}
\newcommand{\changed}[1]{\textcolor{red}{#1}}

@@

\author{M Landreman}
\email[]{mattland@umd.edu}
%\homepage[]{Your web page}
%\thanks{}
\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
\author{H M Smith}
\affiliation{Max-Planck-Institut f\"{u}r Plasmaphysik, 17491 Greifswald, Germany}
\author{A Moll\'{e}n}
\affiliation{Department of Applied Physics, Nuclear Engineering,
  Chalmers University of Technology and Euratom-VR Association,
  G\"oteborg,
  Sweden}
\author{P Helander}
\affiliation{Max-Planck-Institut f\"{u}r Plasmaphysik, 17491 Greifswald, Germany}

@@

To make the problem linear, we make use of the fact that the electrostatic potential is nearly a flux function.
We define $\Phi_0 = \left<\Phi\right>$ and $\Phi_1 = \Phi-\Phi_0$.
We assume $\Phi_1 \ll \Phi_0$,
and we will show shortly that this assumption is self-consistent.
Since $e \Phi_0/T_a \sim 1$ in the drift ordering, then $e \Phi_1/T_a \ll 1$.
We do not expand in the ion charge $Z_a$.
Equation (\ref{eq:Maxwellian}) then gives
$f_{a0} \approx F_{a}\left[ 1 - Z_a e \Phi_1/T_a\right]$ where
\begin{equation}
F_{a}=n_a(\psi) \left[\frac{m_a}{2\pi T_a(\psi)}\right]^{3/2}\exp\left(-\frac{m_a v^2}{2 T_a(\psi)}\right)
\end{equation}
and $n_a = \eta_a \exp(-Z_a e \Phi_0/T_a)$ is the leading order density.
We define the leading-order total energy
$W_{a0} =v^2/2+ Z_a e\Phi_0 /m_a$,
and leading-order $\vect{E}\times\vect{B}$ drift
$\vEo=(c/B^2)(d\Phi_0/d\psi)\vect{B}\times\nabla\psi$.
As the relative differences between $f_{a0}$ and $F_a$, between $W_a$ and $W_{a0}$, and between $\vE$ and $\vEo$
are all small, we may replace the former quantities with the latter ones in (\ref{eq:kinetic2}).
At the same time, we note
\begin{equation}
\frac{\vE\cdot\nabla\psi}{\vma\cdot\nabla\psi} \sim \frac{1}{\epsilon} \frac{Z_a e \Phi_1}{T_a}
\label{eq:radialExB}
\end{equation}
where $\epsilon$ is the relative variation of $B$ on a flux surface, and taking the ratio (\ref{eq:radialExB}) to be small, the $\vE\cdot\nabla\psi$
term in (\ref{eq:kinetic2}) may be neglected. Thus, we obtain
\begin{equation}
\left(\vpar \vect{b} + \vEo + \vma\right) \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
-C_{a}
=-(\vma\cdot\nabla\psi) \left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a}.
\label{eq:kinetic3}
\end{equation}
where $C_a$ is now the collision operator linearized about $F_a$ rather than $f_{a0}$,
\begin{equation}
\left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
=\left[ \frac{1}{p_a}\frac{dp_a}{d\psi} + \frac{Z_a e}{T_a} \frac{d\Phi_0}{d\psi} + \left( x_a^2-\frac{5}{2}\right) \frac{1}{T_a} \frac{dT_a}{d\psi}\right]F_a,
\end{equation}
and $x_a = v/v_a$.
If $F_a$ and $\Phi_0$ are considered known, then (\ref{eq:kinetic3}) is now linear in the unknowns $f_{a1}$,
and $\Phi_1$ has decoupled from the kinetic equations.

@@

For numerical computations, it is convenient to use coordinates for which the ranges of allowed values
are independent of the other coordinates.  As $W_{a0}$ and $\mu$ do not have this property, it is convenient
to switch to coordinates $x_a$ and $\xi = \vpar/v$.  Carrying out this change of variables on the first term
of (\ref{eq:kinetic3}), we find
\begin{equation}
\dot{\vect{r}} \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
=
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a},
\end{equation}
where $\vect{r}$ denotes the position vector,
\begin{equation}
\dot{\vect{r}} = \vpar \vect{b} + \vEo + \vma,
\label{eq:rdot1}
\end{equation}
\begin{equation}
\dot{x}_a = (\vma\cdot\nabla\psi) \left(-\frac{x_a}{2T_a}\frac{dT_a}{d\psi} - \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi}\right),
\label{eq:xdot1}
\end{equation}
and
\begin{equation}
\dot{\xi}_a = -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
-\frac{1-\xi^2}{2 B \xi} \vma\cdot\nabla B
.
\label{eq:xidot1}
\end{equation}
For the rest of this work, we will neglect the $\vma$ term in (\ref{eq:rdot1}), the $dT_a/d\psi$ term
in (\ref{eq:xdot1}), and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1}),
for several reasons.
First, if the  $\vma$ term in (\ref{eq:rdot1}) were retained, we would need to solve a 5D rather
than 4D problem due to the radial coupling (i.e. $\psi$ appearing
as a derivative rather than merely as a parameter). Second,
once radial coupling is dropped, we must also drop the $dT_a/d\psi$ term
in (\ref{eq:xdot1}) and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1})
in order for $\mu$ to be conserved.
Third, dropping these terms conveniently eliminates all dependence of the transport matrix (defined in section \ref{sec:transportMatrix})
on $d T_a/d\psi$, $dB/d\psi$,
\changed{
and $\rho_*$.
Fourth, dropping these terms amounts to taking the limit $\rho_* \to 0$
(while keeping the $d\Phi_0/d\psi$ terms finite), 
and this limit is already complicated and interesting to explore without the extra complexity
of finite-$\rho_*$ corrections.
Fifth, 
we wish to focus on the effects of the radial electric field.
The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.
Finally,
these omitted terms would significantly complicate the analysis in section \ref{sec:sources},
in which we will examine moments of the kinetic equation.
}

Our kinetic equation then becomes
\begin{equation}
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a}
-C_{a}
=-(\vma\cdot\nabla\psi) \left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a}.
\label{eq:kinetic4}
\end{equation}
where the effective particle trajectory equations are
\begin{eqnarray}
\label{eq:fullTrajectories}
\dot{\vect{r}} &=& \vpar \vect{b} + \frac{c}{B^2}\frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi,
\\
\dot{x}_a &=& -(\vma\cdot\nabla\psi) \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi},
\nonumber \\
\dot{\xi}_a &=& -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
.\nonumber
\end{eqnarray}
We will refer to (\ref{eq:fullTrajectories}) as the ``full trajectories.''

@@

\changed{
To summarize, the sources and extra constraint equations serve two independent purposes.
First, when $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints then keep the linear system square (number of equations = number of unknowns) upon discretizatiton.
Second, even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate the null space in the kinetic equation, and the source terms are a convenient way to keep the linear system square upon discretization.
The first problem can be solved with either the source (\ref{eq:sources}) or a general speed-dependent source $S_a(x_a)$.
However, to solve the second problem, the number of constraints should match the dimensionality of the null space.
For this reason, we apply the source (\ref{eq:sources}) with 2 constraints when the Fokker-Planck operator is used,
while we apply the general speed-dependent source $S_a(x_a)$ with $N_x$ constraints when either of the other two collision operators is used.
}

@@

\changed{
We close this section by noting another an important difference between the trajectory models
related to the parallel momentum equations.
Consider the case of \emph{quasisymmetry}, which is the condition \cite{PerAndreiPRL} that
$\vect{B}\times\nabla\psi\cdot\nabla B = A(\psi) \vect{B}\cdot\nabla B$ for some flux function
$A(\psi)$.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
in the trajectories (but retained in the $\partial F_a/\partial \psi$ drive term in (\ref{eq:kinetic4})), the radial neoclassical
current vanishes for all values of $d \Phi_0/d\psi$
if and only if the flux surface is quasisymmetric. This property of quasisymmetric flux surfaces
is known as intrinsic ambipolarity.
Here, we show that intrinsic ambipolarity persists in quasisymmetric geometry
when the $E_r$ terms are retained in the full trajectory
drift-kinetic equation, but not for the DKES or partial trajectory kinetic equations.
This result follows from the $-\sum_a Z_a \left< \int d^3v\, A v_{||} / \Omega_a (\ldots)\right>$
moment of the kinetic equations, i.e. a spatially weighted average of the parallel momentum
moment. For the full trajectories, 
(\ref{eq:fullViscosity}) vanishes in this spatial average, leaving
\begin{equation}
\sum_a Z_a \left< \int d^3v\, f_{a1} \vma\cdot\nabla\psi\right>=0,
\label{eq:noCurrent}
\end{equation}
meaning there is no radial current. However, for the DKES and partial trajectory models, the
spatial average does not annihilate the $d\Phi_0/d\psi$ term,
leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
and therefore the radial current is generally nonzero.  
Consequently, the full trajectory model is the only one of the models that preserves
intrinsic ambipolarity in quasisymmetry for $E_r \ne 0$.
Notice that when the full trajectory model is applied in quasisymmetry,
intrinsic ambipolarity means (\ref{eq:unphysicalEnergyConservation2}) is satisfied even when $d\Phi_0/d\psi \ne 0$,
so the net heat source vanishes for any radial electric field.
}

@@

\begin{figure}[h!]
% Figure generated using m20140219_02_PlotSFINCS2SpeciesEScanForPaper_withSources.m
% ----------------------------------------
% Previous version of figure generated using m20131212_02_PlotSFINCS2SpeciesEScanForPaper.m
% SFINCS run on hopper and edison in 
% /global/scratch2/sd/landrema/20131204-02-sfincs-fortran-hopper-intel
\includegraphics[height=7.5in]{m20140219_02_PlotSFINCS2SpeciesEScanForPaper_withSources.eps}
%\includegraphics{m20131212_02_PlotSFINCS2SpeciesEScanForPaper.eps}
\caption{(Color online)
Fluxes (A)-(B), flow (C), and bootstrap current (D)
computed for a scenario of steep pressure gradient near the edge of W7-X.
Magenta dotted line is the ambipolar $E_r$, effectively identical for the three
trajectory models.
\changed{
The ion particle and heat sources in (D)-(E) are normalized
by a gyro-Bohm transport time $t_{gB}$.}
\label{fig:2species}}
\end{figure}

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
%\newcommand{\tens}[1]{\overleftrightarrow{#1}}
%\newcommand{\tens}[1]{\overleftrightarrow{\vect{#1}}}
\newcommand{\tens}[1]{\vect{#1}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\vi}{v_{\mathrm{i}}}
\newcommand{\me}{m_{\mathrm{e}}}
\newcommand{\mi}{m_{\mathrm{i}}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{\mathrm{ii}}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\todo}[1]{\textcolor{red}{To do: #1}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\Erres}{E_r^{\mathrm{res}}}
\newcommand{\Bpol}{B_{\mathrm{pol}}}
\newcommand{\Sap}{S_{a\mathrm{p}}}
\newcommand{\Sah}{S_{a\mathrm{h}}}
%\newcommand{\changed}[1]{#1}
\newcommand{\changed}[1]{\textcolor{red}{#1}}

@@

\author{M Landreman}
\email[]{mattland@umd.edu}
%\homepage[]{Your web page}
%\thanks{}
\affiliation{Institute for Research in Electronics and Applied Physics, University of Maryland, College Park, MD, 20742, USA}
\author{H M Smith}
\affiliation{Max-Planck-Institut f\"{u}r Plasmaphysik, 17491 Greifswald, Germany}
\author{A Moll\'{e}n}
\affiliation{Department of Applied Physics, %Nuclear Engineering,
  Chalmers University of Technology, %and Euratom-VR Association,
  G\"oteborg,
  Sweden}
\author{P Helander}
\affiliation{Max-Planck-Institut f\"{u}r Plasmaphysik, 17491 Greifswald, Germany}

@@

To make the problem linear, we make use of the fact that the electrostatic potential is nearly a flux function.
We define $\Phi_0 = \left<\Phi\right>$ and $\Phi_1 = \Phi-\Phi_0$.
We assume $\Phi_1 \ll \Phi_0$,
and we will show shortly that this assumption is self-consistent.
Since $e \Phi_0/T_a \sim 1$ in the drift ordering, then $e \Phi_1/T_a \ll 1$.
We do not expand in the ion charge $Z_a$.
Equation (\ref{eq:Maxwellian}) then gives
$f_{a0} \approx F_{a}\left[ 1 - Z_a e \Phi_1/T_a\right]$ where
\begin{equation}
F_{a}=n_a(\psi) \left[\frac{m_a}{2\pi T_a(\psi)}\right]^{3/2}\exp\left(-\frac{m_a v^2}{2 T_a(\psi)}\right)
\end{equation}
and $n_a = \eta_a \exp(-Z_a e \Phi_0/T_a)$ is the leading order density.
We define the leading-order total energy
$W_{a0} =v^2/2+ Z_a e\Phi_0 /m_a$,
and leading-order $\vect{E}\times\vect{B}$ drift
$\vEo=(c/B^2)(d\Phi_0/d\psi)\vect{B}\times\nabla\psi$.
As the relative differences between $f_{a0}$ and $F_a$, between $W_a$ and $W_{a0}$, and between $\vE$ and $\vEo$
are all small, we may replace the former quantities with the latter ones in (\ref{eq:kinetic2}).
At the same time, we note
\begin{equation}
\frac{\vE\cdot\nabla\psi}{\vma\cdot\nabla\psi} \sim \frac{1}{\epsilon} \frac{Z_a e \Phi_1}{T_a}
\label{eq:radialExB}
\end{equation}
where $\epsilon$ is the relative variation of $B$ on a flux surface, and taking the ratio (\ref{eq:radialExB}) to be small, the $\vE\cdot\nabla\psi$
term in (\ref{eq:kinetic2}) may be neglected. Thus, we obtain
\begin{equation}
\left(\vpar \vect{b} + \vEo + \vma\right) \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
-C_{a}
=-(\vma\cdot\nabla\psi) \left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a},
\label{eq:kinetic3}
\end{equation}
where $C_a$ is now the collision operator linearized about $F_a$ rather than $f_{a0}$,
\begin{equation}
\left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
=\left[ \frac{1}{p_a}\frac{dp_a}{d\psi} + \frac{Z_a e}{T_a} \frac{d\Phi_0}{d\psi} + \left( x_a^2-\frac{5}{2}\right) \frac{1}{T_a} \frac{dT_a}{d\psi}\right]F_a,
\end{equation}
and $x_a = v/v_a$.
If $F_a$ and $\Phi_0$ are considered known, then (\ref{eq:kinetic3}) is now linear in the unknowns $f_{a1}$,
and $\Phi_1$ has decoupled from the kinetic equations.

@@

For numerical computations, it is convenient to use coordinates for which the ranges of allowed values
are independent of the other coordinates.  As $W_{a0}$ and $\mu$ do not have this property, it is convenient
to switch to coordinates $x_a$ and $\xi = \vpar/v$.  Carrying out this change of variables on the first term
of (\ref{eq:kinetic3}), we find
\begin{equation}
\dot{\vect{r}} \cdot\left(\nabla f_{a1}\right)_{W_{a0},\mu}
=
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a},
\end{equation}
where $\vect{r}$ denotes the position vector,
\begin{equation}
\dot{\vect{r}} = \vpar \vect{b} + \vEo + \vma,
\label{eq:rdot1}
\end{equation}
\begin{equation}
\dot{x}_a = (\vma\cdot\nabla\psi) \left(-\frac{x_a}{2T_a}\frac{dT_a}{d\psi} - \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi}\right),
\label{eq:xdot1}
\end{equation}
and
\begin{equation}
\dot{\xi}_a = -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
-\frac{1-\xi^2}{2 B \xi} \vma\cdot\nabla B
.
\label{eq:xidot1}
\end{equation}
For the rest of this work, we will neglect the $\vma$ term in (\ref{eq:rdot1}), the $dT_a/d\psi$ term
in (\ref{eq:xdot1}), and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1}),
for several reasons.
First, if the  $\vma$ term in (\ref{eq:rdot1}) was retained, we would need to solve a 5D rather
than 4D problem due to the radial coupling (i.e. $\psi$ appearing
as a derivative rather than merely as a parameter). Second,
once radial coupling is dropped, we must also drop the $dT_a/d\psi$ term
in (\ref{eq:xdot1}) and the $\vma\cdot\nabla B$ term in (\ref{eq:xidot1})
in order for $\mu$ to be conserved.
Third, dropping these terms conveniently eliminates all dependence of the transport matrix (defined in section \ref{sec:transportMatrix})
on $d T_a/d\psi$, $dB/d\psi$,
\changed{
and $\rho_*$.
Fourth, dropping these terms amounts to taking the limit $\rho_* \to 0$
(while keeping the $d\Phi_0/d\psi$ terms finite), 
and this limit is already complicated and interesting to explore without the extra complexity
of finite-$\rho_*$ corrections.
Fifth, 
we wish to focus on the effects of the radial electric field.
The omitted terms may be important in other situations, but here our primary interest is the treatment of the $d\Phi_0/d\psi$ terms.
Finally,
these omitted terms would significantly complicate the analysis in section \ref{sec:sources},
in which we will examine moments of the kinetic equation.
}

Our kinetic equation then becomes
\begin{equation}
\dot{\vect{r}}\cdot \left(\nabla f_{a1}\right)_{x_a,\xi}
+\dot{x}_a \left(\frac{\partial f_{a1}}{\partial x_a}\right)_{\vect{r},\xi}
+\dot{\xi}_a \left(\frac{\partial f_{a1}}{\partial \xi}\right)_{\vect{r},x_a}
-C_{a}
=-(\vma\cdot\nabla\psi) \left(\frac{\partial F_{a}}{\partial \psi}\right)_{W_{a0}}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} F_{a},
\label{eq:kinetic4}
\end{equation}
where the effective particle trajectory equations are
\begin{eqnarray}
\label{eq:fullTrajectories}
\dot{\vect{r}} &=& \vpar \vect{b} + \frac{c}{B^2}\frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi,
\\
\dot{x}_a &=& -(\vma\cdot\nabla\psi) \frac{Z_a e}{2T_a x_a}\frac{d\Phi_0}{d\psi},
\nonumber \\
\dot{\xi}_a &=& -\frac{1-\xi^2}{2 B \xi} \vpar \vect{b}\cdot\nabla B
+\xi(1-\xi^2) \frac{c}{2 B^3} \frac{d\Phi_0}{d\psi}\vect{B}\times\nabla\psi\cdot\nabla B
.\nonumber
\end{eqnarray}
We will refer to (\ref{eq:fullTrajectories}) as the ``full trajectories.''

@@

\changed{
To summarize, the sources and extra constraint equations serve two independent purposes.
First, when $E_r \ne 0$, the sources are needed to eliminate the conservation problems, and the extra constraints then keep the linear system square (number of equations = number of unknowns) upon discretization.
Second, even when $E_r = 0$, and even for the DKES trajectories in which sources are not required, the constraints are needed to eliminate the null space in the kinetic equation, and the source terms are a convenient way to keep the linear system square upon discretization.
The first problem can be solved with either the source (\ref{eq:sources}) or a general speed-dependent source $S_a(x_a)$.
However, to solve the second problem, the number of constraints should match the dimensionality of the null space.
For this reason, we apply the source (\ref{eq:sources}) with 2 constraints when the Fokker-Planck operator is used,
while we apply the general speed-dependent source $S_a(x_a)$ with $N_x$ constraints when either of the other two collision operators is used.
}

@@

\changed{
We close this section by noting another important difference between the trajectory models
related to the parallel momentum equations.
Consider the case of \emph{quasisymmetry}, which is the condition %\cite{PerAndreiPRL} 
that
$\vect{B}\times\nabla\psi\cdot\nabla B = A(\psi) \vect{B}\cdot\nabla B$ for some flux function
$A(\psi)$ \cite{PerAndreiPRL}.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
in the trajectories (but retained in the $\partial F_a/\partial \psi$ drive term in (\ref{eq:kinetic4})), the radial neoclassical
current vanishes for all values of $d \Phi_0/d\psi$
if and only if the flux surface is quasisymmetric. This property of quasisymmetric flux surfaces
is known as intrinsic ambipolarity.
Here, we show that intrinsic ambipolarity persists in quasisymmetric geometry
when the $E_r$ terms are retained in the full trajectory
drift-kinetic equation, but not for the DKES or partial trajectory kinetic equations.
This result follows from the $-\sum_a Z_a \left< \int d^3v\, A v_{||} / \Omega_a (\ldots)\right>$
moment of the kinetic equations, i.e. a spatially weighted average of the parallel momentum
moment. For the full trajectories, 
(\ref{eq:fullViscosity}) vanishes in this spatial average, leaving
\begin{equation}
\sum_a Z_a \left< \int d^3v\, f_{a1} \vma\cdot\nabla\psi\right>=0,
\label{eq:noCurrent}
\end{equation}
meaning there is no radial current. However, for the DKES and partial trajectory models, the
spatial average does not annihilate the $d\Phi_0/d\psi$ term,
leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
and therefore the radial current is generally nonzero.  
Consequently, the full trajectory model is the only one of the models that preserves
intrinsic ambipolarity in quasisymmetry for $E_r \ne 0$.
Notice that when the full trajectory model is applied in quasisymmetry,
intrinsic ambipolarity means (\ref{eq:unphysicalEnergyConservation2}) is satisfied even when $d\Phi_0/d\psi \ne 0$,
so the net heat source vanishes for any radial electric field.
}

@@

\begin{figure}[h!]
% Figure generated using m20140219_02_PlotSFINCS2SpeciesEScanForPaper_withSources.m
% ----------------------------------------
% Previous version of figure generated using m20131212_02_PlotSFINCS2SpeciesEScanForPaper.m
% SFINCS run on hopper and edison in 
% /global/scratch2/sd/landrema/20131204-02-sfincs-fortran-hopper-intel
\includegraphics[height=7.5in]{m20140219_02_PlotSFINCS2SpeciesEScanForPaper_withSources.eps}
%\includegraphics{m20131212_02_PlotSFINCS2SpeciesEScanForPaper.eps}
\caption{(Color online)
Fluxes (A)-(B), flow (C), and bootstrap current (D)
computed for a scenario of steep pressure gradient near the edge of W7-X.
Magenta dotted line is the ambipolar $E_r$, effectively identical for the three
trajectory models.
\changed{
The ion particle and heat sources in (E)-(F) are normalized
by a gyro-Bohm transport time $t_{gB}$.}
\label{fig:2species}}
\end{figure}
